NumLin: Linear Types for Linear Algebra

### 2019-01-11 write-up/paper/intro.tex
Beefed up intro; added a stronger opening to language overview

>>>>

Programmers writing numerical software often find themselves caught on
the horns of a dilemma. The foundational, low-level linear algebra
libraries such as BLAS and LAPACK offer programmers very precise
control over the memory lifetime and usage of vector and matrix
values. However, this power comes paired with the responsibility to
manually manage the memory associated with each array object, and in
addition to bringing in the familiar difficulties of reasoning about
lifetimes, aliasing and sharing that plague low-level systems
programming, this also moves the APIs away from the linear-algebraic,
mathematical style of thinking that numerical programmers want to use.

As a result, programmers often turn to higher-level languages such as
Matlab, R and Numpy, which offer very high level array abstractions
that can be viewed as ordinary mathematical values. This makes
programming safer, as well as making prototyping and verification much
easier, since it lets programmers write programs which bear a closer
resemblance to the formulas that the mathematicians and statisticians
designing these algorithms prefer to work with, and ensures that
program bugs will reflect incorrectly-computed values rather than heap
corruption.

The intention is that these languages can use libraries BLAS and
LAPACK, without having to expose programmers to explicit memory
management. However, this benefit comes at a price: because user
programs do not worry about aliasing, the language implementations
cannot in general exploit the underlying features of the low-level
libraries that let them explicitly manage and reuse memory. As a
result, programs written in high-level statistical languages can
be much less memory-efficient than programs that make full use
of the powers the low-level APIs offer.

So in practice, programmers face a tradeoff: they can eschew safety
and exploit the full power of the underlying linear algebra libraries,
or they can obtain safety at the price of unneeded copies and worse
memory efficiency. In this work, show that this tradeoff is not a
fundamental one.

### 2019-01-11 write-up/paper/lang_and_examples.tex
Beefed up intro; added a stronger opening to language overview

Linearity is at the heart of \lang. Linearity allows us to express a
pure-functional API for numerical library routines that mutate arrays and
matrices. Linearity also restricts aliasing of (values which represent)
pointers.

>>>>

The core type theory of \lang\ is a nearly off-the-shelf linear type
theory, supporting familiar features such as linear function spaces
$A \multimap B$ and tensor products $A \otimes B$. We adopt linearity
-- the restriction that each program variable be used at most once --
since it allows us to express purely functional APIs for numerical
library routines that mutate arrays and matrices~\cite{Wadler90}. Due
to linearity, values cannot alias and are only used once, which means
that linearly-typed updates result in no \emph{observable} mutation.

As a result, programmers can reason about \lang\ expressions as if they
were ordinary mathematical expressions -- as indeed they are! We are
merely adopting a stricter type discipline than usual to make managing
memory safe.
